---
title: "Decision Modeling for Agroecology & Conservation"
author: "Cory Whitney"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(decisionSupport)
library(ggplot2)
library(learnr)
library(shiny)
library(DiagrammeR)

source("functions/monte_carlo.R")
source("functions/apples_sheep_in_class.R")
source("functions/apple_sheep_pareto.R")

```

# **Introduction**

Welcome to the interactive tutorial on **Decision Modeling for Agroecology & Conservation**. 
This tutorial will guide you through:

- **Theoretical foundations** of decision modeling 
- **Identifying decision options and outcomes** 
- **Structuring causal models** using Directed Acyclic Graphs (DAGs)
- **Monte Carlo simulation** for decision-making
- **Expert-elicited probabilities** and model validation
- **Evaluating models** with EVPI and Pareto-optimality

---

## **1. The Need for Decision Modeling in Agroecology**

- **Application of decision modeling in agroecology**
- **The role of expert knowledge**
- **Decision structuring** 

```{r question1, echo=FALSE}
quiz(
  question("What is a good reason to use a probabilistic approach to support decision-making in agroecology?",
    answer("It captures uncertainty in outcomes and trade-offs.", correct = TRUE),
    answer("It provides fixed deterministic solutions.", correct = FALSE),
    answer("It replaces the need for expert knowledge.", correct = FALSE),
    answer("It avoids the need for validation.", correct = FALSE)
  )
)
```

---

## **2. Identifying Decision Options and Outcomes**

**Activity:** Identify and list **decision options** and **outcomes of interest** from the perspective of decision-makers.

```{r decision_options, echo=FALSE}
textInput("decision_options", "List the decision options relevant to an agroecological problem:", placeholder="Example: Adopting agroforestry, expanding irrigation, switching to organic inputs...")
```

```{r decision_outcomes, echo=FALSE}
textInput("decision_outcomes", "List the key outcomes of interest for decision-makers:", placeholder="Example: Yield improvement, biodiversity conservation, financial returns...")
```

---

## **3. Structuring a Causal Decision Model (DAGs)**

**Activity:** Sketch a decision model using the interactive diagram tool or upload a hand-drawn image.

### **Option 1: Interactive DAG Drawing**

Share a link of a graph you build in [Loopy](https://ncase.me/loopy/) by [Explorable Explanations](http://explorableexplanations.com/)

### **Option 2: Upload a Hand-Drawn Model**

Sketch out a decision model on a piece of paper and share with us.

```{r upload_model, echo=FALSE}
fileInput("model_upload", "Upload your decision model (optional)", accept = c("image/png", "image/jpeg"))
```

---

## **4. Implementing the Apple-Sheep Monte Carlo Model**

Now for a very simple **Monte Carlo Simulation** in R. We will analyze a **decision**. The choice of a farmer to start agroforestry (apple trees + sheep) or maintain a treeless sheep system. Use `set.seed` and change the number of simulations to 1000.

```{r apples_model, exercise=TRUE}

# set.seed(42)
num_simulations <- 100

# Define ranges per ha per year Euro
# lower and upper
apple_income <- runif(n = num_simulations, 
                      min = 3000, 
                      max = 60000)

apple_costs <- runif(n = num_simulations, 
                      min = 15000, 
                     max = 30000)

apple_profits <- apple_income - apple_costs
```

```{r apples_model-solution}

set.seed(42)
num_simulations <- 1000

# Define ranges per ha per year Euro
# lower and upper
apple_income <- runif(n = num_simulations, 
                      min = 3000, 
                      max = 60000)

apple_costs <- runif(n = num_simulations, 
                      min = 15000, 
                     max = 30000)

apple_profits <- apple_income - apple_costs
```

Add sheep to the horticulture system. So we can judge the relative benefits of transition to a silvopastoral agroforestry system. Add the sheep and apple profits as a last step in the function.

```{r sheep_model, exercise=TRUE}

# Euro per ha per year
sheep_income <- runif(n = num_simulations, 
                      min = 2000, 
                      max = 5000)

sheep_costs <- runif(n = num_simulations, 
                     min = 1000, 
                     max = 2500)

sheep_profits <- sheep_income - sheep_costs
```

```{r sheep_model-solution}

# Euro per ha per year
sheep_income <- runif(n = num_simulations, 
                      min = 2000, 
                      max = 5000)

sheep_costs <- runif(n = num_simulations, 
                     min = 1000, 
                     max = 2500)

sheep_profits <- sheep_income - sheep_costs

total_profits <- apple_profits + sheep_profits
```

Plot a histogram of the profits. Change the color of the apple_profits to grey. 

```{r plot_profits_overlay, exercise=TRUE}
hist(total_profits, col = "white")
hist(apple_profits, add = TRUE, col = "pink")
hist(sheep_profits, add = TRUE, col = "black")
```

```{r plot_profits_overlay-solution}
hist(total_profits, col = "white")
hist(apple_profits, add = TRUE, col = "grey")
hist(sheep_profits, add = TRUE, col = "black")
```

---

## **5. Pareto-Optimality)**

Now we will calcualte PAreto-Optimality. We will use the concept of the Pareto front to find the best trade-offs between two competing objectives. We can learn which solutions are *optimal* (better in at least one way, without being worse in another) and which solutions are *dominated* (we could improve at least one objective without making the other worse).

First we need to increase the model complexity a little by adding management options. We add stocking density (sheep per hectare), tree density (trees per hectare), proportion of tree area, and proportion of sheep area (1 - proportion_tree_area). Adjust these numbers to reflect your best estimate.

```{r management_variables, exercise=TRUE}

# Define management variables per hectare
max_stocking_density <- 20  # Max sheep per ha before overgrazing effects occur
stocking_density <- runif(num_simulations, min = 5, max = max_stocking_density)

max_tree_density <- 500   # Maximum trees per hectare
tree_density <- runif(num_simulations, min = 50, max = max_tree_density)

# Define proportion of tree area vs. sheep area
proportion_tree_area <- runif(num_simulations, min = 0.2, max = 0.8)
proportion_sheep_area <- 1 - proportion_tree_area  # Remaining area for sheep
```

```{r management_variables-solution}

# Define management variables per hectare
max_stocking_density <- 40  # Max sheep per ha before overgrazing effects occur
stocking_density <- runif(num_simulations, min = 5, max = max_stocking_density)

max_tree_density <- 200   # Maximum trees per hectare
tree_density <- runif(num_simulations, min = 50, max = max_tree_density)

# Define proportion of tree area vs. sheep area
proportion_tree_area <- runif(num_simulations, min = 0.2, max = 0.8)
proportion_sheep_area <- 1 - proportion_tree_area  # Remaining area for sheep

total_profits_agroforestry <- apple_profits + (sheep_profits * proportion_sheep_area)
total_profits_treeless <- sheep_profits  # No tree component in treeless system
```

Adjust these biodiversity numbers to reflect your best estimate. For example, depending on the ecosystem, each tree likely adds far fewer species and each sheep likely has far less of a negative impact. 

```{r biodiversity_variables, exercise=TRUE}

# Define species richness and grazing impact factors per hectare
species_richness_per_tree <- 0.05  # Each tree adds 0.05 species per ha
species_loss_per_sheep <- 0.02  # Each sheep reduces 0.02 species per ha

# Calculate biodiversity per ha
tree_species_richness <- tree_density * species_richness_per_tree
grazing_species_loss <- stocking_density * species_loss_per_sheep

# Compute biodiversity outcomes per ha
biodiversity_agroforestry <- tree_species_richness - grazing_species_loss
biodiversity_treeless <- -grazing_species_loss  # No trees, only grazing effect

# Calculate difference (decision impact of trees)
# measures the financial difference when switching to trees
total_profits <- total_profits_agroforestry - total_profits_treeless

# Calculate difference (decision impact of trees on biodiversity)
# measures the net gain/loss in biodiversity from switching to trees
biodiversity_impact <- biodiversity_agroforestry - biodiversity_treeless

# Define a weighting factor for biodiversity impact (convert biodiversity into € equivalent)
biodiversity_value_per_species <- 500  # Assume each additional species is worth €500

# Compute a combined utility function
utility <- total_profits + (biodiversity_impact * biodiversity_value_per_species)
```

```{r biodiversity_variables-solution}

# Define species richness and grazing impact factors per hectare
species_richness_per_tree <- 0.01  # Each tree adds 0.001 species per ha
species_loss_per_sheep <- 0.001  # Each sheep reduces 0.001 species per ha

# Calculate biodiversity per ha
tree_species_richness <- tree_density * species_richness_per_tree
grazing_species_loss <- stocking_density * species_loss_per_sheep

# Compute biodiversity outcomes per ha
biodiversity_agroforestry <- tree_species_richness - grazing_species_loss
biodiversity_treeless <- -grazing_species_loss  # No trees, only grazing effect

# Calculate difference (decision impact of trees)
# measures the financial difference when switching to trees
total_profits <- total_profits_agroforestry - total_profits_treeless

# Calculate difference (decision impact of trees on biodiversity)
# measures the net gain/loss in biodiversity from switching to trees
biodiversity_impact <- biodiversity_agroforestry - biodiversity_treeless

# Define a weighting factor for biodiversity impact (convert biodiversity into € equivalent)
biodiversity_value_per_species <- 500  # Assume each additional species is worth €500

# Compute a combined utility function
utility <- total_profits + (biodiversity_impact * biodiversity_value_per_species)
```

For the Pareto front plot we will first add the data together, store results in dataframe and plot a scatter plot. Change the color of the points in the scatter plot to green. 

```{r results_data, exercise=TRUE}

results <- data.frame(
  total_profits,
  biodiversity_impact
)

# Scatter plot
plot(results$biodiversity_impact, results$total_profits,
     col = "blue", pch = 19, xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")
```

```{r results_data-solution}

results <- data.frame(
  total_profits,
  biodiversity_impact
)

# Scatter plot
plot(results$biodiversity_impact, results$total_profits,
     col = "green", pch = 19, xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")
```

Now we can look for the non-dominated points where no other solution is better in both biodiversity and total profits. A point (biodiversity, profit) is Pareto-optimal if no other point has higher biodiversity and higher profits. Change the color of the points to blue.

```{r pareto_points, exercise=TRUE}

# Sort results by biodiversity impact
results <- results[order(results$biodiversity_impact, decreasing = TRUE),]

# Initialize Pareto front
pareto_front <- results[1, ]  # Start with the first point

# Loop through sorted results and keep only non-dominated points
for (i in 2:nrow(results)) {
  if (results$total_profits[i] > tail(pareto_front$total_profits, 1)) {
    pareto_front <- rbind(pareto_front, results[i, ])
  }
}

# Plot all points
plot(results$biodiversity_impact, results$total_profits,
     col = "green", pch = 19, xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")

# Highlight Pareto front points
points(pareto_front$biodiversity_impact, pareto_front$total_profits,
       col = "red", pch = 19, type = "o", lwd = 2)
```

```{r pareto_points-solution}

# Sort results by biodiversity impact
results <- results[order(results$biodiversity_impact, decreasing = TRUE),]

# Initialize Pareto front
pareto_front <- results[1, ]  # Start with the first point

# Loop through sorted results and keep only non-dominated points
for (i in 2:nrow(results)) {
  if (results$total_profits[i] > tail(pareto_front$total_profits, 1)) {
    pareto_front <- rbind(pareto_front, results[i, ])
  }
}

# Plot all points
plot(results$biodiversity_impact, results$total_profits,
     col = "green", pch = 19, xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")

# Highlight Pareto front points
points(pareto_front$biodiversity_impact, pareto_front$total_profits,
       col = "blue", pch = 19, type = "o", lwd = 2)
```

Now we can zoom in on the Pareto-optimal points and examine the conditions for key operational variables we calculated earlier.

```{r pareto_conditions, exercise=TRUE}

# Extract operational conditions for Pareto-optimal points
pareto_conditions <- results[results$biodiversity_impact %in% pareto_front$biodiversity_impact &
                               results$total_profits %in% pareto_front$total_profits,]

# Merge with original dataset to include operational variables
pareto_conditions <- merge(pareto_conditions, data.frame(
  biodiversity_impact, total_profits, stocking_density, tree_density, proportion_tree_area),
  by = c("biodiversity_impact", "total_profits"))

# Show Pareto-optimal conditions
print(pareto_conditions)
```

```{r pareto_conditions-solution}

# Extract operational conditions for Pareto-optimal points
pareto_conditions <- results[results$biodiversity_impact %in% pareto_front$biodiversity_impact &
                               results$total_profits %in% pareto_front$total_profits,]

# Merge with original dataset to include operational variables
pareto_conditions <- merge(pareto_conditions, data.frame(
  biodiversity_impact, total_profits, stocking_density, tree_density, proportion_tree_area),
  by = c("biodiversity_impact", "total_profits"))

# Show Pareto-optimal conditions
print(pareto_conditions)
```

Finally we can plot the Pareto. Make the color of the points blue.

```{r plot_pareto, exercise=TRUE}
# Determine dynamic position based on plot coordinates
text_positions <- ifelse(
  pareto_front$biodiversity_impact > median(pareto_front$biodiversity_impact), 2, 4
)

# Adjust for very high or low points
text_positions <- ifelse(
  pareto_front$total_profits > quantile(pareto_front$total_profits, 0.75), 1, text_positions
)
text_positions <- ifelse(
  pareto_front$total_profits < quantile(pareto_front$total_profits, 0.25), 3, text_positions
)

# Plot Pareto front
plot(pareto_front$biodiversity_impact, pareto_front$total_profits,
     col = "red", pch = 19, type = "o", lwd = 2,
     xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")

# Add footnote at the bottom
mtext("S = Stocking Density (sheep/ha), T = Tree Density (trees/ha), P = Proportion of Tree Area", 
      side = 1, line = 4, cex = 0.8)

# Create multi-line labels with line breaks
labels <- paste0("S:", round(pareto_conditions$stocking_density, 1), "\n",
                 "T:", round(pareto_conditions$tree_density, 1), "\n",
                 "P:", round(pareto_conditions$proportion_tree_area, 2))

# Annotate with dynamically positioned text
text(pareto_front$biodiversity_impact, pareto_front$total_profits,
     labels = labels, pos = text_positions, cex = 0.8)
```

```{r plot_pareto-solution}
# Determine dynamic position based on plot coordinates
text_positions <- ifelse(
  pareto_front$biodiversity_impact > median(pareto_front$biodiversity_impact), 2, 4
)

# Adjust for very high or low points
text_positions <- ifelse(
  pareto_front$total_profits > quantile(pareto_front$total_profits, 0.75), 1, text_positions
)
text_positions <- ifelse(
  pareto_front$total_profits < quantile(pareto_front$total_profits, 0.25), 3, text_positions
)

# Plot Pareto front
plot(pareto_front$biodiversity_impact, pareto_front$total_profits,
     col = "blue", pch = 19, type = "o", lwd = 2,
     xlab = "Biodiversity Impact (Species Change per ha)",
     ylab = "Total Profits (€ per ha)", main = "Pareto Front: Agroforestry vs. Treeless System")

# Add footnote at the bottom
mtext("S = Stocking Density (sheep/ha), T = Tree Density (trees/ha), P = Proportion of Tree Area", 
      side = 1, line = 4, cex = 0.8)

# Create multi-line labels with line breaks
labels <- paste0("S:", round(pareto_conditions$stocking_density, 1), "\n",
                 "T:", round(pareto_conditions$tree_density, 1), "\n",
                 "P:", round(pareto_conditions$proportion_tree_area, 2))

# Annotate with dynamically positioned text
text(pareto_front$biodiversity_impact, pareto_front$total_profits,
     labels = labels, pos = text_positions, cex = 0.8)
```

---

## **7. Final Reflection**

- What insights did you gain from this tutorial?
- How would you improve your decision model?
- How does this relate to **real-world agroecology and conservation?**

```{r feedback, echo=FALSE}
textInput("feedback", "Share your key takeaways:", placeholder="Write your reflections here...")
```

---

## **Next Steps**

- Apply Monte Carlo simulations to your own research
- Continue refining expert-elicited models

```{r outro_message, echo=FALSE}
cat("Thanks for participating in this tutorial! 🎉")
```



