---
title: "Decision Modeling for Agroecology"
author: "Hoa Do, Cory Whitney"
date: "`r Sys.Date()`"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(decisionSupport)
library(ggplot2)
library(learnr)
library(shiny)
library(DiagrammeR)
source("AF_Maize_Script.R")
```

## NLKH: xen canh xoai (mango) và ngo (maize)

```{r set_seed, exercise=TRUE}
n <- 100
set.seed(123) 
```

```{r AF_mango_yield, exercise=TRUE}
#NLKH: xen canh xoai (mango) và ngo (maize)
AF_mango_yield <- runif(n, 6000, 10000) # giá trị nhỏ nhất và lớn nhất (kg/ha)
```

```{r AF_maize_yield, exercise=TRUE}
AF_maize_yield <- runif( n, 3000, 5000) # (kg/ha)
AF_maize_yield
```



```{r AF_cost, exercise=TRUE}
AF_cost <- runif(n, 15, 30) # (trieu VND/ha)
```

#Ngo doc canh

```{r mono_maize_yield, exercise=TRUE}
mono_maize_yield <- runif (n, 4000, 8000) # (ton/ha)
```

```{r maize_cost, exercise=TRUE}
maize_cost <- runif(n, 7, 15) # 10 trieu VND/ha
```

```{r maize_price, exercise=TRUE}
maize_price <- runif(n, 0.003, 0.007) # trieu VND/kg
```

```{r mango_price, exercise=TRUE}
mango_price <- runif(n, 0.002, 0.007) # trieu VND/kg
```


# 2) Rủi ro hạn hán

```{r p_drought, exercise=TRUE}

p_drought <- runif(n, 0.10, 0.50)     # mô phỏng xác suất hạn hán 
```



```{r drought, exercise=TRUE}

drought <- rbinom(n, size = 1, prob = p_drought)  # 1 = năm hạn hán
```




```{r loss_maize, exercise=TRUE}

loss_maize <- ifelse(drought == 1, runif(n, 0.30, 0.70), 0)
```




```{r loss_mango, exercise=TRUE}

loss_mango <- ifelse(drought == 1, runif(n, 0.10, 0.50), 0)
```



```{r AF_mango_yield_drought, exercise=TRUE}

AF_mango_yield_drought <- AF_mango_yield * (1 - loss_mango)
```




```{r AF_maize_yield_drought, exercise=TRUE}

AF_maize_yield_drought <- AF_maize_yield * (1 - loss_maize)
```


```{r mono_maize_yield_drought, exercise=TRUE}

mono_maize_yield_drought <- mono_maize_yield * (1 - loss_maize)
mono_maize_yield_drought
```


# --- Lợi nhuận ---

```{r AF_profit_drought, exercise=TRUE}
AF_profit_drought <- (AF_mango_yield_drought * mango_price + AF_maize_yield_drought * maize_price) - AF_cost
mean(AF_profit_drought > 0)
hist(
  AF_profit_drought,
  breaks = 40,
  col = "lightblue",
  border = "white",
  main = "Nông lâm kết hợp",
  ylab = "Tần suất",
  xlab = "Lợi nhuận (triệu VND/ha)"
)
```



# Kết quả kỳ vọng của quyết định

```{r figure3, exercise=TRUE}
mono_profit_drought <- (mono_maize_yield_drought * maize_price) - maize_cost
mean(mono_profit_drought > 0)
hist(
  mono_profit_drought,
  breaks = 40,
  col = "rosybrown1",
  border = "white",
  main = "Ngô độc canh",
  ylab = "Tần suất",
  xlab = "Lợi nhuận (triệu VND/ha)"
)
```


Thanks for following this tutorial. As thanks, here is an open space for running your scripts. 

```{r blank, exercise=TRUE}

```

---

## References

Do H, Luedeling E, Whitney C. Decision analysis of agroforestry options reveals adoption risks for resource-poor farmers. Agron Sustain Dev. 2020;40(3):20. doi:10.1007/s13593-020-00624-5

Fernandez E, Do H, Luedeling E, Luu TTG, Whitney C. Prioritizing farm management interventions to improve climate change adaptation and mitigation outcomes—a case study for banana plantations. Agron Sustain Dev. 2022;42(4):76. doi:10.1007/s13593-022-00809-0

Do H, Whitney C, La N, Storm H, Luedeling E. Adapting agroforestry to upland farming systems: narratives from smallholder farmers in Northwest Vietnam. Agron Sustain Dev. 2024;44(2):17. doi:10.1007/s13593-024-00954-8

Do H, Whitney C, Storm H, Nguyen H, La N, Luedeling E. Markets and incentives strongly drive agroforestry adoption: insights from ethnic minority smallholders in Son La, Vietnam. Agroforestry Systems. Published online 2025:99:129. doi:10.1007/s10457-025-01199-4

Whitney, C, K Shepherd, and E Luedeling. “Decision Analysis Methods Guide; Agricultural Policy for Nutrition.” World Agroforestry (ICRAF) Working Paper series, no. 275 (2018): 40. http://dx.doi.org/10.5716/WP18001.PDF.

Whitney, Cory, Denis Lanzanova, Caroline Muchiri, Keith D. Shepherd, Todd S. Rosenstock, Michael Krawinkel, John R. S. Tabuti, and Eike Luedeling. “Probabilistic Decision Tools for Determining Impacts of Agricultural Development Policy on Household Nutrition.” Earth’s Future 6, no. 3 (2018): 359–72. https://doi.org/10.1002/2017EF000765.

