## **3. Structuring a Causal Decision Model (DAGs)**

**Activity:** Sketch a decision model using the interactive diagram tool or upload a hand-drawn image.

### **Option 1: Interactive DAG Drawing**

Share a link of a graph you build in [Loopy](https://ncase.me/loopy/) by [Explorable Explanations](http://explorableexplanations.com/)

### **Option 2: Upload a Hand-Drawn Model**

Sketch out a decision model on a piece of paper and share with us.

```{r upload_model, echo=FALSE}
fileInput("model_upload", "Upload your decision model (optional)", accept = c("image/png", "image/jpeg"))
```

Or you can adjust this one if you are familiar with DOT language. 

```{r decision_model, exercise=TRUE}
DiagrammeR::grViz("
digraph decision_model {
  # Define node styles
  node [shape=ellipse, style=filled, fillcolor=lightyellow] Apple_Income Sheep_Income Total_incomes ;
  node [shape=ellipse, style=filled, fillcolor=lightpink] Sheep_cost Apple_cost Total_costs;
  node [shape=ellipse, style=filled, fillcolor=lightblue] Outcome_Profit;
  
  # Add layout
  graph [rankdir=LR]
  
  # Add labels
  Apple_Income [label = 'Apple Income (€)']
  Sheep_Income [label = 'Sheep Income (€)']
  Total_incomes [label = 'Total Income']
  Apple_cost [label = 'Apple Cost (€)']
  Sheep_cost [label = 'Sheep Cost (€)']
  Total_costs [label = 'Total Cost']
  Outcome_Profit [label = 'Net Profit (€)']

  # Define relationships
  Sheep_cost -> Total_costs;
  Apple_cost -> Total_costs;
  Apple_Income -> Total_incomes;
  Sheep_Income -> Total_incomes;
  Total_costs -> Outcome_Profit;
  Total_incomes -> Outcome_Profit
}
")

```

---
