## **4. Implementing a Monte Carlo Model**

Now for a very simple **Monte Carlo Simulation** in R. We will analyze a **decision**. The choice of a farmer to start agroforestry (apple trees + sheep) or maintain a treeless sheep system. Use `set.seed` and change the number of simulations to 1000. 

```{r apples_model, exercise=TRUE}

# set.seed(42)
num_simulations <- 100

# Define ranges per ha per year Euro
# lower and upper
apple_income <- runif(n = num_simulations, 
                      min = 3000, 
                      max = 60000)

apple_costs <- runif(n = num_simulations, 
                      min = 15000, 
                     max = 30000)

apple_profits <- apple_income - apple_costs
```

```{r apples_model-solution}

set.seed(42)
num_simulations <- 1000

# Define ranges per ha per year Euro
# lower and upper
apple_income <- runif(n = num_simulations, 
                      min = 3000, 
                      max = 60000)

apple_costs <- runif(n = num_simulations, 
                      min = 15000, 
                     max = 30000)

apple_profits <- apple_income - apple_costs
```

Add sheep to the horticulture system. So we can judge the relative benefits of transition to a silvopastoral agroforestry system. Add the sheep and apple profits as a last step in the function.

```{r sheep_model, exercise=TRUE}

# Euro per ha per year
sheep_income <- runif(n = num_simulations, 
                      min = 2000, 
                      max = 5000)

sheep_costs <- runif(n = num_simulations, 
                     min = 1000, 
                     max = 2500)

sheep_profits <- sheep_income - sheep_costs
```

```{r sheep_model-solution}

# Euro per ha per year
sheep_income <- runif(n = num_simulations, 
                      min = 2000, 
                      max = 5000)

sheep_costs <- runif(n = num_simulations, 
                     min = 1000, 
                     max = 2500)

sheep_profits <- sheep_income - sheep_costs

total_profits <- apple_profits + sheep_profits
```

Plot a histogram of the profits. Change the color of the apple_profits to grey. 

```{r plot_profits_overlay, exercise=TRUE}
hist(total_profits, col = "white")
hist(apple_profits, add = TRUE, col = "pink")
hist(sheep_profits, add = TRUE, col = "black")
```

```{r plot_profits_overlay-solution}
hist(total_profits, col = "white")
hist(apple_profits, add = TRUE, col = "grey")
hist(sheep_profits, add = TRUE, col = "black")
```

```{r question2, echo=FALSE}
quiz(
  question("What does the histogram tell us about uncertainty?",
    answer("There is high variation in profits, suggesting high uncertainty.", correct = TRUE),
    answer("All outcomes are nearly the same, meaning low uncertainty.", correct = FALSE)
  )
)
```

---
