## **5. Monte Carlo Simulation with `decisionSupport`**

Now we use the `decisionSupport` package for a structured Monte Carlo simulation. This approach uses probability distributions for inputs and a decision model function to compute Net Present Value (NPV) outcomes, giving us a principled way to compare the agroforestry and treeless systems.

### **Define Input Estimates**

Adjust the lower bounds to reflect realistic minimum values.

```{r apple_sheep_estimates, exercise=TRUE}
input_estimates <- data.frame(
  variable = c("sheep_income", "sheep_cost", "apple_income", "apple_cost", "discount_rate"),
  lower = c(3000, 1000, 30000, 15000, 10),
  median = NA,
  upper = c(5000, 2500, 60000, 30000, 10),
  distribution = c("posnorm", "posnorm", "posnorm", "posnorm", "const"),
  label = c("Income from sheep (euro/year)", "Cost of sheep (euro/year)",
            "Income from apple (euro/year)", "Cost of apple (euro/year)", "Discount Rate")
)
```

```{r apple_sheep_estimates-solution}
input_estimates <- data.frame(
  variable = c("sheep_income", "sheep_cost", "apple_income", "apple_cost", "discount_rate"),
  lower = c(300, 100, 3000, 1500, 1),
  median = NA,
  upper = c(5000, 2500, 60000, 30000, 10),
  distribution = c("posnorm", "posnorm", "posnorm", "posnorm", "const"),
  label = c("Income from sheep (euro/year)", "Cost of sheep (euro/year)",
            "Income from apple (euro/year)", "Cost of apple (euro/year)", "Discount Rate")
)
```

### **Build the Decision Model Function**

Adjust as you see fit. Add the income from agroforestry `AF_income` to the return for example.

```{r apple_sheep_model_function, exercise=TRUE}
model_function <- function() {
  AF_income <- sheep_income + apple_income
  AF_cost <- sheep_cost + apple_cost
  AF_final_result <- AF_income - AF_cost
  sheep_only <- sheep_income - sheep_cost
  Decision_benefit <- AF_final_result - sheep_only

  AF_NPV <- discount(AF_final_result, discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_sheep_only <- discount(sheep_only, discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_decision <- discount(Decision_benefit, discount_rate = discount_rate, calculate_NPV = TRUE)

  return(list(
    NPV_Agroforestry_System = AF_NPV,
    NPV_Treeless_System = NPV_sheep_only,
    NPV_decision = NPV_decision
  ))
}
```

```{r apple_sheep_model_function-solution}
model_function <- function() {
  AF_income <- sheep_income + apple_income
  AF_cost <- sheep_cost + apple_cost
  AF_final_result <- AF_income - AF_cost
  sheep_only <- sheep_income - sheep_cost
  Decision_benefit <- AF_final_result - sheep_only

  AF_NPV <- discount(AF_final_result, discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_sheep_only <- discount(sheep_only, discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_decision <- discount(Decision_benefit, discount_rate = discount_rate, calculate_NPV = TRUE)

  return(list(
    NPV_Agroforestry_System = AF_NPV,
    NPV_Treeless_System = NPV_sheep_only,
    NPV_decision = NPV_decision,
    AF_income = AF_income
  ))
}
```

### **Run the Monte Carlo Simulation**

Adjust the number of simulations to `1000`.

```{r apple_sheep_mc, exercise=TRUE}
apple_sheep_mc_simulation <- mcSimulation(
  estimate = as.estimate(input_estimates),
  model_function = model_function,
  numberOfModelRuns = 100,
  functionSyntax = "plainNames"
)
```

```{r apple_sheep_mc-solution}
apple_sheep_mc_simulation <- mcSimulation(
  estimate = as.estimate(input_estimates),
  model_function = model_function,
  numberOfModelRuns = 1000,
  functionSyntax = "plainNames"
)
```

### **Plot the Results**

```{r plot_mc_results, exercise=TRUE}
plot_distributions(mcSimulation_object = apple_sheep_mc_simulation,
                   vars = c("NPV_Agroforestry_System", "NPV_Treeless_System", "NPV_decision"))
```

### **Expected Value of Perfect Information (EVPI)**

The EVPI tells us how much we would benefit from having perfect information about each uncertain variable.

```{r evpi_calc, exercise=TRUE}
mcSimulation_table <- data.frame(apple_sheep_mc_simulation$x, apple_sheep_mc_simulation$y[3])
evpi <- multi_EVPI(mc = mcSimulation_table, first_out_var = "NPV_decision")
evpi
```

---
