## **6. The Reference Model: `monte_carlo.R`**

`monte_carlo.R` is the **foundational model**. It defines five uncertain variables, a simple decision model function, and a single call to `mcSimulation()`. All inputs are treated as *annual* values — there is no explicit time horizon. The `discount()` function is applied to a single number, so it converts an annual net benefit into a present value using the discount rate.

### **Workflow**

1. Build an `input_estimates` data frame with probability distributions for each variable.
2. Write a `model_function()` that computes outcomes from those variables.
3. Call `mcSimulation()` to sample from the distributions and run the model thousands of times.
4. Plot and interpret the resulting distributions.

### **Step 1: Define the input estimates**

The table below covers exactly five variables. `posnorm` draws from a positive normal distribution bounded by `lower` and `upper`. `const` fixes the variable at its `lower` value with no uncertainty.

```{r mc_basic_estimates, exercise=TRUE}
input_estimates <- data.frame(
  variable = c("sheep_income", "sheep_cost", "apple_income", "apple_cost", "discount_rate"),
  lower = c(3000, 1000, 30000, 15000, 10),
  median = NA,
  upper = c(5000, 2500, 60000, 30000, 10),
  distribution = c("posnorm", "posnorm", "posnorm", "posnorm", "const"),
  label = c("Income from sheep (euro/year)", "Cost of sheep (euro/year)",
            "Income from apple (euro/year)", "Cost of apple (euro/year)", "Discount Rate")
)
```

### **Step 2: Write the model function**

Each call to `model_function()` receives one draw of each variable from the distributions above. The function computes the net outcome for the agroforestry system, the treeless sheep system, and the difference (the decision benefit). `discount()` converts each annual result into a Net Present Value (NPV).

```{r mc_basic_model, exercise=TRUE}
model_function <- function() {
  AF_income  <- sheep_income + apple_income
  AF_cost    <- sheep_cost  + apple_cost
  AF_final_result  <- AF_income  - AF_cost
  sheep_only       <- sheep_income - sheep_cost
  Decision_benefit <- AF_final_result - sheep_only

  AF_NPV        <- discount(AF_final_result,  discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_sheep_only <- discount(sheep_only,       discount_rate = discount_rate, calculate_NPV = TRUE)
  NPV_decision   <- discount(Decision_benefit, discount_rate = discount_rate, calculate_NPV = TRUE)

  return(list(
    NPV_Agroforestry_System = AF_NPV,
    NPV_Treeless_System     = NPV_sheep_only,
    NPV_decision            = NPV_decision
  ))
}
```

### **Step 3: Run `mcSimulation()`**

Increase `numberOfModelRuns` to 1000 and observe how the output distributions change.

```{r mc_basic_sim, exercise=TRUE}
apple_sheep_mc <- mcSimulation(
  estimate          = as.estimate(input_estimates),
  model_function    = model_function,
  numberOfModelRuns = 100,
  functionSyntax    = "plainNames"
)
plot_distributions(mcSimulation_object = apple_sheep_mc,
                   vars = c("NPV_Agroforestry_System", "NPV_Treeless_System", "NPV_decision"))
```

```{r mc_basic_sim-solution}
apple_sheep_mc <- mcSimulation(
  estimate          = as.estimate(input_estimates),
  model_function    = model_function,
  numberOfModelRuns = 1000,
  functionSyntax    = "plainNames"
)
plot_distributions(mcSimulation_object = apple_sheep_mc,
                   vars = c("NPV_Agroforestry_System", "NPV_Treeless_System", "NPV_decision"))
```

```{r question_mc_basic, echo=FALSE}
quiz(
  question("In this basic model, what does `discount()` receive as its first argument?",
    answer("A single annual net benefit value.", correct = TRUE),
    answer("A vector of yearly benefits over a time horizon.", correct = FALSE),
    answer("The cumulative sum of all simulation runs.", correct = FALSE)
  )
)
```

---
